---
title: "Week 5 Advanced Graphics Practical"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      eval = TRUE)
```

# Instructions

## Before you start

This practical will be run using *posit.cloud* to avoid individual machine
troubleshooting. You will find it helpful to have several resources close at
hand: The lecture slides and [the ggplot2 Book](https://ggplot2-book.org/).
The R help will be indispensable for completing this exercise, remember for any 
`function()` you can type ?function to quickly pull up the help page.

## The goal

This practical will allow you to practice making basic data visualisations using
R.

## Learning objectives
1. Use ggplot2 to make a density plot.
2. Use ggplot2 to make a time series plot with annotations
3. Add error bars to observations in ggplot2
3. Use ggplot2 to plot multivariate data with a stacked bar plot
4. Use ggplot2 to plot multivariate data with a heat map
5. Use ggplot2 to plot multivariate data with a bubble chart
6. Use ggplot2 to produce annotated maps

## The way it's going to be run.

Write your own `R`-code and run it. That is the best way to learn how this
program works. Also, you have an excellent Teaching Assistant resource - ask
them or the instructor the questions you have.

::: {.alert .alert-info}
The text in the blue box marked as **Task (number):** states what you need to do
:::

```{r}
# The code in the window marked like this shows where you need to write your code
#
# 
# If I ask you to assess the result or answer a question, you should write the 
#text here as a comment starting with two hashes (##)
```

## Assessment.

Submit your `R-markdown` file via BrightSpace - **The assessment is a Pass/Fail
based on how you write and annotate the code - You need to show us you know what
you are doing and NOT copying someone else's code.**

# Tasks

## Importing Danish Covid-19 data
We will use the same Danish Covid-19 dataset that we used in the lecture. Import
this dataset into a data frame called `covid`. Keep in mind that for this file:

* Columns are separated by “;” (semicolon)
* Decimals are commas
* The first line contains header names

::: {.alert .alert-info}
**Task 1a:** Import the data from the file `Test_pos_over_time.csv` and place it
in a data frame called `covid`. Then format the dates correctly.
:::

```{r}
#Import the data
covid <- -----(----)
#Format the dates correctly
covid$---- <- ----(----)
```

## Histograms and density plots

Now that the Covid-19 dataset is loaded, imagine that you want to get an
impression of the distribution for the number of new positive cases on any given
day during the pandemic. You need a histogram and a density plot.

::: {.alert .alert-info}
**Task 1b:** Using ggplot2, draw a histogram of the variable `NewPositive` from the dataframe `covid`.
:::

```{r}
#Draw a histogram
----(----)
----(----) + ----()
```

You will notice that most days have very few positive cases, but some
days have many cases, making it difficult to observe a lot of variation
in the cases. Change the y axis to a log10 scale using the function
`scale_y_log10()` to better observe the high numers of positive tests
per day. Reflect on how you might have distorted the data
visualisation, taking principle 3 of making good plots into account
from the lecture.

::: {.alert .alert-info}
**Task 1c:** Using ggplot2, draw a histogram of the variable
`NewPositive` from the dataframe `covid` with a log10 scaled y axis.
How might this have interfered with principles for good graphics?
:::

```{r}
#Draw a histogram with a log10 y axis
----(----) + ----() +
  ----()
#How might this log scalehave interfered with principles for good #graphics?
##The log axis may mislead the viewer, as it looks like there was less
##variation in positive test numbers than there actually was. However,
##it also makes the few days with high test numbers more visible. So
##there is a tradeoff here which may or may not make sense depending on
##the purpose of the data visualisation.
```

Other than histograms, you have also learnt another way of showing data
distributions called density plots. In ggplot2 this can be achieved with a 
geom called `geom_density`. Make a density plot as an extra layer to
your histogram, and change the colour of the line so that it stands
out.

::: {.alert .alert-info}
**Task 1d:** Using ggplot2, draw a density plot of the variable
`NewPositive` from the dataframe `covid`.
:::

```{r}
#Draw a density plot of the variable `NewPositive` from the dataframe
#`covid`.
----(----) + ----()
```

## Annotating a time series

Let's assume that we want to make a plot showing how much Covid-19 PCR testing was carried out in Denmark at different times in the pandemic.

::: {.alert .alert-info}
**Task 2a:** Using ggplot2, draw a time series plot with the date on the x axis
and the number of tests carried out (variable name `Tested`) on the y axis. Give
your plot axes helpful labels. Use a line plot.
:::

```{r}
#Draw a time series plot with the date on the x axis and the number of tests
#carried out (variable name `Tested`) on the y axis.
----(----) + ----() +
----(----)
```

Imagine you want to compare the Danish testing regime with that carried out in
other countries, so you decide to plot the number of tests carried out divided
by the total Danish population, which we can assume is 5.8 million for the whole
period.

::: {.alert .alert-info}
**Task 2b:** Using ggplot2, draw a time series plot with the date on the x axis
and the number of tests carried out per million people (variable name `Tested`)
on the y axis. Give your plot axes helpful labels. Use a line plot.
:::

```{r}
#Draw a time series plot with the date on the x axis and the number of tests
#carried out (variable name `Tested`) on the y axis.
----(----) + ----() +
----(----)
```

Graphics like these can become very powerful if annotated with some explanatory
data. For example, we can indicate certain important dates that occurred during
the pandemic to help make sense of the testing numbers. You can add vertical
lines to a plot using `geom_vline()`, and text using `geom_text()`.

::: {.alert .alert-info}
**Task 2c:** Using ggplot2, draw a time series plot with the date on the x axis
and the number of tests carried out per million people (variable name `Tested`)
on the y axis. Give your plot axes helpful labels, use a line plot. Also add
transparent red vertical lines for Christmas Eve 2020 and 2021, and transparent
blue vertical lines for New Year's Eve 2020 and 2021. Label these lines with
blue and red text.
:::

```{r}
#Plot Covid-19 tests per day per million people with Christmas and New Year's
#labeled for 2020 and 2021 as described above.
----(----) + ----() ...
```

Vertical lines are good for annotating specific points in time, but what about 
if you want to annotate a longer period in time? A transparent rectangle could
be useful here. Let's say you wanted to annotate the above plot with a coloured
box in the background indicating the time from the first Danish school
shutdown (11/3/2020) until primary schools were reopened physically (18/5/2020). 

::: {.alert .alert-info}
**Task 2d:** Using ggplot2, draw a time series plot with the date on the x axis
and the number of tests carried out per million people (variable name `Tested`)
on the y axis. Give your plot axes helpful labels, use a line plot. Also add
a coloured box from 11/3/2020 to 18/5/2020 using `geom_rect`, and label it
appropriately using `geom_text`. Note that the four aesthetics you will need to
specify for `geom_rect()` are `xmin`, `xmax`, `ymin`, and `ymax`. The order of
your geoms will be important.
:::

```{r}
#Plot Covid-19 tests per day per million people with the first school shutdown
#period marked with a rectangle as described above.
----(----) +
----(----) + 
----() + ----(----) +
  ----(----)
```

## For loop practice

Imagine that you are a data scientist working for Sundhedsstyrelsen and have
been asked to prepare the following six graphics: a plot of daily positive tests
for each month from the first 6 months of 2022. You decide to produce these graphics
using a for loop to save time. You make three vectors: one containing the start
date for each month, one containing the finish date for each month,
and one containing the name of each month. You then write a for loop that
iterates over these vectors plotting each month one at a time, saving the plot
as a pdf file using the `ggsave()` function. You will probably need to use the
`paste0()` function to make the filename each time (to paste `.pdf` onto the end
of the month name to give each plot file a unique name).

::: {.alert .alert-info}
**Task 3:** Using ggplot2 and a for loop, draw a time series plot for each of
the first six months of 2022 with the date on the x axis and the number of
positive tests on the y axis. Give your plot axes helpful labels and the month
name as a title, use line plots. 
:::

```{r}
#Make three vectors  containing the start date for each month, one containing
#the finish date for each month, and one containing the name of each month.

---- <- ----(c(----))
---- <- ----(c(----))
---- <- c(----)

#Using ggplot2 and a for loop, draw a time series plot for each of
#the first six months of 2022 with the date on the x axis and the number of
#positive tests on the y axis.

for (----) {
----(----) + ----() + 
    ----(----) +
    ----(----) + ----(----)

----(----(----))
}
```

## Error bars

Although error bars can hide the underlying distribution of your data, they are
very common in spite of this shortcoming and it's important to know how to plot
them onto your data when you're using ggplot2.

For this section we will use a set of quantitative PCR data that looks at the 
abundance of sulfate-reducing bacteria beneath the seabed. Each observation has 
several variables including: `Depth` (cm below the seafloor, anegative number), 
`Abundance` (number of dsrB genes), and `SD` (the standard deviation for each 
measurement). First let's import the data from a file called `qPCR_example_dataset.tsv`.

::: {.alert .alert-info}
**Task 4a:** Import the qPCR dataset into a dataframe called `qpcr`. The file
`qPCR_example_dataset.tsv` contains a dataset where:
* Columns are separated by tabs
* The first line is a header
* Decimal points are points '.'
:::

```{r}
#Import the qPCR dataset into a dataframe called `qpcr`
---- <- ----(----)
```

Now we will plot the data. Plot the depth on the y axis and the abundance of
dsrB genes per gram sediment (`Abundance` variable) on the x axis. The standard
deviation (`SD`) of each data point should be plotted as an error bar to the
left and right of each data point using the `geom_errorbar()` function. Use
`geom_point()` and `geom_line(orientation = "y")` to show the data with points
connected by a line ordered by the points orientation on the y axis.

::: {.alert .alert-info}
**Task 4b:** Plot the qPCR dataset as described above. Include axis labels.
:::

```{r}
#Plot the qPCR dataset as described above.
----(----) + ----() +
  ----(----) +
  ----(----) + ----(----) +
  ----(----)
```

## Multivariate plotting

First import the bedroom dust dataset that we used in the lecture. This dataset
contains one column for each bacterial phylum found in a series of 9 bedrooms.
Each row is an observation of one bedroom.

::: {.alert .alert-info}
**Task 5a:** Import the bedroom dust dataset into a dataframe called `br`.
The file `bedroom_dust.tsv` contains a dataset where:
* Columns are separated by tabs
* The first line is a header
* Decimal points are points '.'
* The first column contains the row names
:::

```{r}
#Import the bedroom dust dataset
---- <- ----(-----)
```

Now to prepare this dataset for plotting, we need to
(a) convert the row names to a column
(b) convert it to "tidy" format to condense the 32 phylum variables down 
to two variables, one for the phylum name and the other for the phylum
abundance

::: {.alert .alert-info}
**Task 5b:** Convert the row names (accessible through the function `row.names()`)
to a separate column and use the `pivot_longer()` function from the `tidyr` package 
to condense the dataset to three columns: bedroom name, phylum name, and fractional 
abundance of that phylum.
:::

```{r}
#Make a column for the sample names in the bedroom dust dataframe
---- <- ----(----)
#Make a new dataframe with just three variables with `pivot_longer`.
----(----)
---- <- ----(----)
```

Let's plot the data using a stacked barplot.

::: {.alert .alert-info}
**Task 5c:** Using ggplot2, make a stacked bar plot where the bedroom name is
mapped to the x axis variable, the abundance of each phylum is mapped to the y
axis variable and the phylum is mapped to the fill colour of the bar. Make
sure the bedroom names are legible. What problem do you immediately notice 
with this graphic?
:::

```{r}
#Make a stacked barplot to show the abundance of each phylum present in
#each bedroom

----(----) +
  ----() +
  ----(----)

#What problem do you immediately notice with this graphic?
##There are too many colours to determine which phylum is which.
```

We can make this plot more legible by combining some of the less abundant
phyla together into a category we can call "Other".

::: {.alert .alert-info}
**Task 5d:** Using ggplot2, make a stacked bar plot where the bedroom name is
mapped to the x axis variable, the abundance of each phylum is mapped to the y
axis variable and the phylum is mapped to the fill colour of the bar. Plot
only Actinobacteria, Firmicutes, and Proteobacteria and combine the rest into a
category called "Other" using the `fct_other()` function from the `forcats`
package. Why did we use `fct_other()` here and use `fct_lump_n()` for a similar
operation during the lecture? Which of the 9 principles of good graphics is being
violated here?
:::

```{r}
#Lump the low abundance (not the top 5) phyla together
----(----)
---- <- ----(----)

#Make a stacked bar plot like the last step but with low-abundance phyla
#lumped together
----(----) +
  ----() +
  ----(----)

#Why did we use `fct_other()`/`geom_col()` here and use `fct_lump_n()`/`geom_bar()`
#for a similar operation during the lecture?
##fct_lump_n/geom_bar looks at the frequency of observations with an
##observation (like we had in the invasive species dataset) while 
##fct_other/geom_col works where the abundances are a variable in each
##observation, rather than being calculated from the number of observations.

#Which of the 9 principles of good graphics is being violated here?
##1. Show your data - we are hiding those phyla that are in "Others" 
```

A good alternative to stacked bar plots is the heat map. Let's make one of these
now. Note that you may have edited your dataframe to include the "Other" category
in the last task, so you might need to re-import your data.

::: {.alert .alert-info}
**Task 5e:** Using ggplot2 and `geom_tile()`, make a heatmap where the phyla are
on the y axis and the bedrooms are on the x axis. Use a sensible colour scale.
:::

```{r}
---- <- ----(----)
---- <- ----(----)
---- <- ----(----)

----(----) +
  ----() + 
  ----(----) +
  ----()
```

Another alternative to stacked bar plots heat maps is a bubble chart.
Let's make one of these now, but where we use both size and colour to indicate
the abundance of each bacterial phylum. Make zero values absent.

::: {.alert .alert-info}
**Task 5f:** Using ggplot2 and `geom_point()`, make a bubble chart where the
phyla are on the y axis and the bedrooms are on the x axis, with the size and
colour of each point indicating the fractional abundance of each phylum. If a
phylum is absent from a bedroom measurement (a zero value) then it should be
absent from the plot too.
:::

```{r}
---- <- ----[----]

----(----) +
  ----(----) + 
  ----(----) +
  ----()
```

## Annotating Maps

As we covered in the lecture, ggplot2 can be made to plot maps and annotate them.

::: {.alert .alert-info}
**Task 6a:** Make a map of the world and draw a horizontal dotted red line on the
Tropic of Capricorn (23.43628º South) using `geom_hline()`. Label it using `geom_text()`.
:::

```{r}
----(----)
----(----)

---- <- ----(----)
----(----) + ----() + ----(----) +
  ----(----)
```

::: {.alert .alert-info}
**Task 6b:** Make a map of Europe and draw a red rectangle approximately around Denmark using `geom_rect()`. Label it using `geom_text()`.
:::

```{r}
----(----)
----(----)

----(----) + ----() + 
  ----(----) +
  ----(----) +
  ----(----) +
  ----(----)
```
